---
title: "How Life Expectancy is Impacted by Air Quality"
output: 
  html_document: 
    fig_height: 6
    fig_width: 9
    toc: yes
    toc_depth: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message= FALSE)
source("useful_functions.R")
load("output/analytical_data.RData")
library(mice)
library(sandwich)
library(lmtest)
library(survey)
library(texreg)
library(pander)
library(Amelia)
library(ggplot2)
```

```{r imputation}
#drop after 2014, because all CO2 data missing
wdidata <- as.data.frame(subset(wdidata, time<2015 & counname!="St. Martin (French part)",
                  select=c("counname","time","lifexpec","pm2.5","co2emmission","gdpcap",
                           "gdpcapgro","urbanpop")))
wdidata.impute <- amelia(wdidata, m=5, ts="time", cs="counname")
```

```{r functions, echo=FALSE}

lm_mi <- function(formula, imp) {
  b <- se <- R2 <- NULL
  
  for(i in 1:imp$m) {
    #run the model
    model <- lm(formula, data=imp$imputations[[i]])
    #collect the results
    b <- cbind(b, coef(model))
    se <- cbind(se, summary(model)$coef[,2])
    R2 <- c(R2,summary(model)$r.squared)
  }
  
  #pool the results
  b.pool <- apply(b, 1, mean)
  between.var <- apply(b, 1, var)
  within.var <- apply(se^2, 1, mean)
  se.pool <- sqrt(within.var+between.var+between.var/imp$m) 
  t.pool <- b.pool/se.pool 
  pvalue.pool <- (1-pnorm(abs(t.pool)))*2 
  coefficients <- data.frame(b.pool, se.pool, t.pool, pvalue.pool)
  
  #the mean R2 value
  r.squared <- mean(R2)
  #we can also grap n and p from the last model since 
  #they should be the same across all iterations
  n <- nobs(model)
  p <- length(model$coefficients)-1
  #go ahead and calculate BIC.null
  bic.null <- n*log(1-r.squared)+p*log(n)
  
  #return everything in a list
  return(list(coef=coefficients,
              n=n,
              r.squared=r.squared,
              bic.null=bic.null))
}

convertModel <- function(model) {
  coef <- subset(model$coef, !grepl("counname", rownames(model$coef)))
  tr <- createTexreg(
    coef.names = rownames(coef), 
    coef = coef$b.pool, 
    se =coef$se.pool, 
    pvalues = coef$pvalue.pool,
    gof.names = c("R2","BIC (null)","N"), 
    gof = c(model$r.squared, model$bic.null, model$n), 
    gof.decimal = c(T,F,F)
  )
}
```

# Introduction
<a href="#top">Back to top</a>

```{r figure, echo=FALSE}
ggplot(wdidata, aes(x=co2emmission, y=lifexpec))+
         geom_point()+
         geom_smooth(se=FALSE)+
         geom_smooth(method="lm", aes(group=counname), color="red", se=FALSE)+
         theme_bw()

ggplot(wdidata, aes(x=pm2.5, y=lifexpec))+
         geom_point()+
         geom_smooth(se=FALSE)+
         geom_smooth(method="lm", aes(group=counname), color="red", se=FALSE)+
         theme_bw()

ggplot(wdidata, aes(x=gdpcap, y=lifexpec))+
         geom_point()+
         geom_smooth(se=FALSE)+
         geom_smooth(method="lm", aes(group=counname), color="red", se=FALSE)+
         theme_bw()

ggplot(wdidata, aes(x=gdpcapgro, y=lifexpec))+
         geom_point()+
         geom_smooth(se=FALSE)+
         geom_smooth(method="lm", aes(group=counname), color="red", se=FALSE)+
         theme_bw()

ggplot(wdidata, aes(x=urbanpop, y=lifexpec))+
         geom_point()+
         geom_smooth(se=FALSE)+
         geom_smooth(method="lm", aes(group=counname), color="red", se=FALSE)+
         theme_bw()

#synthetic scale
ggplot(wdidata, aes(x=gdpcap, y=co2emmission))+
  geom_point()+
  geom_smooth(se=FALSE)+
  geom_smooth(method="lm", aes(group=counname), color="red", se=FALSE)+
  theme_bw()

wdidata$bad_air <- scale(scale(wdidata$pm2.5)+scale(wdidata$co2emmission))

ggplot(wdidata, aes(x=bad_air, y=lifexpec))+
         geom_point()+
         geom_smooth(se=FALSE)+
         geom_smooth(method="lm", aes(group=counname), color="red", se=FALSE)+
         theme_bw()


wdidata$pm2.5_gdpcap <- scale(scale(wdidata$pm2.5)+scale(wdidata$gdpcap))

ggplot(wdidata, aes(x=pm2.5_gdpcap, y=lifexpec))+
         geom_point()+
         geom_smooth(se=FALSE)+
         geom_smooth(method="lm", aes(group=counname), color="red", se=FALSE)+
         theme_bw()

wdidata$co2_gdpcap <- scale(scale(wdidata$co2emmission)+scale(wdidata$gdpcap))

ggplot(wdidata, aes(x=co2_gdpcap, y=lifexpec))+
         geom_point()+
         geom_smooth(se=FALSE)+
         geom_smooth(method="lm", aes(group=counname), color="red", se=FALSE)+
         theme_bw()

```

```{r models, results='asis'}
model1 <- lm_mi(log(lifexpec)~log(co2emmission)+counname, wdidata.impute)
model2 <- lm_mi(log(lifexpec)~log(co2emmission)+log(gdpcap)+counname+as.factor(time), wdidata.impute)
model3 <- lm_mi(log(lifexpec)~log(co2emmission)+log(pm2.5)+log(gdpcap)+counname+as.factor(time), wdidata.impute)
model4 <- lm_mi(log(lifexpec)~log(co2emmission)+log(gdpcapgro)+log(pm2.5)+log(gdpcap)+counname+as.factor(time), wdidata.impute)
model5 <- lm_mi(log(lifexpec)~log(co2emmission)+log(urbanpop)+log(gdpcapgro)+log(pm2.5)+log(gdpcap)+counname+as.factor(time), wdidata.impute)
htmlreg(lapply(list(model1, model2, model3, model4, model5), convertModel), digits=4)

```

```{r models-n-log, results='asis'}
model1 <- lm_mi(lifexpec~co2emmission+counname, wdidata.impute)
model2 <- lm_mi(lifexpec~co2emmission+I(gdpcap/1000)+counname+as.factor(time), wdidata.impute)
model3 <- lm_mi(lifexpec~co2emmission+pm2.5+I(gdpcap/1000)+counname+as.factor(time), wdidata.impute)
model4 <- lm_mi(lifexpec~co2emmission+gdpcapgro+pm2.5+I(gdpcap/1000)+counname+as.factor(time), wdidata.impute)
model5 <- lm_mi(lifexpec~co2emmission+urbanpop+gdpcapgro+pm2.5+I(gdpcap/1000)+counname+as.factor(time), wdidata.impute)
htmlreg(lapply(list(model1, model2, model3, model4, model5), convertModel), digits=4)
```


```{r models-n-log-switch, results='asis'}
model1 <- lm_mi(lifexpec~pm2.5+counname, wdidata.impute)
model2 <- lm_mi(lifexpec~pm2.5+I(gdpcap/1000)+counname+as.factor(time), wdidata.impute)
model3 <- lm_mi(lifexpec~pm2.5+co2emmission+I(gdpcap/1000)+counname+as.factor(time), wdidata.impute)
model4 <- lm_mi(lifexpec~pm2.5+gdpcapgro+co2emmission+I(gdpcap/1000)+counname+as.factor(time), wdidata.impute)
model5 <- lm_mi(lifexpec~pm2.5+urbanpop+gdpcapgro+co2emmission+I(gdpcap/1000)+counname+as.factor(time), wdidata.impute)
htmlreg(lapply(list(model1, model2, model3, model4, model5), convertModel), digits=4)
```



Use this R Markdown to perform the main analysis for the project. I use this basically as a lab notebook. It contains the main analysis and a variety of sensitivity analysis. The code in this documents serves as a baseline for the eventual tables and figures that will go into the paper. At the same time it will serve as a record of all supplementary analyses performed. 